---
title: "HMM fitting and visualization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HMM fitting and visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  message = FALSE,
  warning = FALSE
)
```

This vignette walks through the Hidden Markov Model (HMM) workflow in FlyDreamR, including fitting and plotting.

## Fit an HMM

After preparing your dataset with `HMMDataPrep()`, you can fit an HMM using:

- `HMMbehavr()` (single-threaded)
- `HMMbehavrFast()` (parallel; faster on multi-core machines)

```{r}
# Load demo data
meta_file <- system.file("extdata", "Metadata_Monitor1.csv", package = "FlyDreamR")
data_dir <- system.file("extdata", package = "FlyDreamR")
dt <- HMMDataPrep(
  metafile_path = meta_file,
  result_dir    = data_dir,
  ldcyc         = 12,
  day_range     = c(1, 2)
)

# Serial
res <- HMMbehavr(behavtbl = dt, it = 100, ldcyc = 12)

# Parallel
res_prl <- HMMbehavrFast(behavtbl = dt, it = 100, ldcyc = 12, n_cores = 4)
```

## Plot HMM state profiles

FlyDreamR provides convenient plotting helpers:

- `HMMplot()` for heatmap hypnograms
- `HMMFacetedPlot()` to facet individual hypnograms
- `HMMSinglePlot()` to save individual hypnograms to disk

```{r}
# Heatmap hypnograms
HMMplot(res)

# Use a different built-in palette
HMMplot(res, color_palette = "AG")

# Compare serial vs parallel fits
library(patchwork)
HMMplot(res, color_palette = "AG") / HMMplot(res_prl, color_palette = "AG")

# User-defined colors
HMMplot(
  res,
  color_palette = "user",
  user_colors = c(
    "State0" = "#FF0060",
    "State1" = "#F6FA70",
    "State2" = "#00DFA2",
    "State3" = "#0079FF"
  )
)

# Faceted + save-to-disk helpers
HMMFacetedPlot(res)
HMMSinglePlot(res)
```

## Summarize time spent in each state

The HMM results typically include:

- `TimeSpentInEachState` (summaries)
- `VITERBIDecodedProfile` (decoded state sequences)

You can visualize time spent per state with `ggplot2`:

```{r}
ggplot2::ggplot(res$TimeSpentInEachState,
                ggplot2::aes(x = state_name, y = time_spent, fill = state_name)) +
  ggplot2::geom_boxplot(width = 0.35, outliers = FALSE) +
  ggplot2::facet_grid(day ~ phase, scales = "free_y") +
  ggplot2::theme_minimal()
```
