---
title: "Analyzing Sleep States with FlyDreamR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analyzing Sleep States with FlyDreamR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  message = FALSE,
  warning = FALSE
)
```

`FlyDreamR` is an R package for analyzing *Drosophila* Activity Monitor (DAM) data, with workflows for:

-   Preparing DAM master/monitor files for downstream analyses
-   Traditional sleep metrics (sleep time, bout statistics, activity indices, brief awakenings)
-   Hidden Markov Models (HMMs) to infer discrete behavioral states and visualize state dynamics
-   An optional Shiny app for interactive analysis

This vignette gives a short roadmap and points you to the more focused vignettes included with the package.

## What to read next

-   **Getting data into FlyDreamR**: `vignette("data-prep", package = "FlyDreamR")`
-   **Traditional sleep analyses**: `vignette("traditional-sleep", package = "FlyDreamR")`
-   **HMM fitting and visualization**: `vignette("hmm-workflow", package = "FlyDreamR")`
-   **Running the Shiny app**: `vignette("shiny-app", package = "FlyDreamR")`

## Installation (typical options)

If you have a source tarball (e.g., `FlyDreamR_1.0.0.tar.gz`), you can install from source:

```{r}
install.packages(c("devtools", "ggplot2", "ggetho", "magrittr", "patchwork"))

d <- tempdir()
untar("FlyDreamR_1.0.0.tar.gz", compressed = "gzip", exdir = d)
devtools::install(file.path(d, "FlyDreamR"),
                  dependencies = TRUE,
                  repos = "https://cloud.r-project.org/")
```

Then load the package:

```{r}
library(FlyDreamR)
```

## Getting help

-   Check documentation: `?HMMbehavr`
-   View examples: `example(HMMplot)`
-   Report issues / request features: add your preferred tracker link in `DESCRIPTION` (e.g., `URL` / `BugReports`)

## Citation

If you use FlyDreamR in your research, please cite:

```         
Ghosh, A. & Harbison, S.T. (2026). Inferring the genetic basis of sleep states in Drosophila melanogaster using Hidden Markov models. GitHub repository.
```
