---
title: "Analyzing Sleep States with FlyDreamR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analyzing Sleep States with FlyDreamR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  message = FALSE,
  warning = FALSE
)
```

`FlyDreamR` is an R package for analyzing *Drosophila* Activity Monitor (DAM) data, with workflows for:

-   Preparing DAM master/monitor files for downstream analyses
-   Traditional sleep metrics (sleep time, bout statistics, activity indices, brief awakenings)
-   Hidden Markov Models (HMMs) to infer discrete behavioral states and visualize state dynamics
-   An optional Shiny app for interactive analysis

This vignette gives a short roadmap and points you to the more focused vignettes included with the package.

## What to read next

-   **Getting data into FlyDreamR**: `vignette("data-prep", package = "FlyDreamR")`
-   **Traditional sleep analyses**: `vignette("traditional-sleep", package = "FlyDreamR")`
-   **HMM fitting and visualization**: `vignette("hmm-workflow", package = "FlyDreamR")`
-   **Running the Shiny app**: `vignette("shiny-app", package = "FlyDreamR")`

## Installation (typical options)

From GitHub (recommended)

```{r}
# Install from GitHub (devtools or remotes)
install.packages(c('devtools','remotes'), repos='https://cloud.r-project.org')
remotes::install_github('orijitghosh/FlyDreamR', upgrade = 'never')

# Pin to a specific release
remotes::install_github('orijitghosh/FlyDreamR@v1.0.0', upgrade = 'never')
```

Then load the package:

```{r}
library(FlyDreamR)
```

## Getting help

-   Check documentation: `?HMMbehavr`
-   View examples: `example(HMMplot)`

## Citation

If you use FlyDreamR in your research, please cite:

```         
Ghosh, A. & Harbison, S.T. (2026). Inferring the genetic basis of sleep states in Drosophila melanogaster using Hidden Markov models. GitHub repository.
```
