---
title: "Running the FlyDreamR Shiny app"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running the FlyDreamR Shiny app}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  message = FALSE,
  warning = FALSE
)
```

FlyDreamR can be used entirely from the console, but it also ships with a Shiny app for interactive analysis.

## Launch the app

If the Shiny app directory is installed with the package, you can usually locate it via `system.file()`.
(If you keep the app outside the package, point `runApp()` at that folder instead.)

```{r}
# Preferred (if bundled in the package)
app_dir <- system.file("Shiny-FlyDreamR", package = "FlyDreamR")
shiny::runApp(app_dir)
```

In the original demo workflow, the app was launched from a local folder path:

```{r}
shiny::runApp("PATH/TO/APP/Shiny-FlyDreamR_1.0.0/")
```

## Notes

- The first launch may install additional packages (depending on how the app is configured).
- For intramural/shared use, the app can also be deployed to a Shiny Server / Posit Connect environment.
